Estimating demographic bias on tests of children's early vocabulary

### 2022-09-07 paper/manuscript.Rmd
adding APA manuscript for topiCS submission

---
title: "Estimating demographic bias on tests of children's early vocabulary"
bibliography: library.bib
csl: apa6.csl
document-params: "10pt, letterpaper"

author-information: > 
    \author{{\large \bf George Kachergis, Nathan Francis, and Michael C. Frank}
    \AND {\large \tt  \{kachergis, nathan99, mcfrank\} @stanford.edu} \\ Department of Psychology, Stanford Unviersity \\ Stanford, CA 94305 USA
    }

abstract: >
    Children's early language skill has been linked to later educational outcomes, making
    it important to measure early language accurately. Parent-reported instruments such as the 
    Communicative Development Inventories (CDIs) have been shown to provide reliable and valid
    measures of children's aggregate early language skill. However, CDIs contain
    hundreds of vocabulary items, some of which may not be heard (and thus learned) equally
    often by children of varying backgrounds. This study used a database of American English 
    CDIs to identify words demonstrating strong bias for particular demographic groups of children, on 
    dimensions of sex (male vs. female), race (white vs. non-white), and maternal education 
    (high vs. low). For each dimension, we identified dozens of strongly biased items, and showed that 
    eliminating even some of these items can reduce the magnitude of differences between groups. 
    Additionally, we investigated how well the relative frequency of words spoken to young girls vs.
    boys predicted sex-based word learning bias, and discuss possible sources of demographic
    differences in early word learning.

keywords: >
    language acquisition; word learning; measuring instrument bias; demographic bias

output: cogsci2016::cogsci_paper
final-submission: \cogscifinalcopy
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=3, fig.height=3, fig.crop = F, 
                      fig.pos = "tb", fig.path='figs/',
                      echo=F, warning=F, cache=TRUE, 
                      message=F, sanitize = T)
```

@@

# Introduction

@@

We focus here on the MacArthur-Bates Communicative Development Inventories [CDIs; @fenson1994; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
CDIs have shown good predictive validity [e.g., @fenson1994; @bornstein2012stability; @duff2015infant].
Our primary focus is the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.
For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.
Children's total vocabulary score on the CDI:WS is tightly correlated with other facets of early language (e.g., grammatical competence and gesture), suggesting that the language system is "tightly woven" [@frank2021].
Due to these desirable properties, CDIs have been adapted to dozens of languages, and a central repository of CDI data contributed from all over the world has been created [Wordbank; @frank2017; @frank2021].

<!-- [^1]: The CDI: Words and Gestures form includes a subset of the CDI:WS items and targets children 12 to 18 months of age, measuring both comprehension and production. -->

@@

The outline of this paper is as follows.
First, we introduce the Wordbank data and the IRT model, and examine the overall size of demographic differences in early word learning based on the full CDI:WS.
We then fit the IRT model to both groups for each demographic factor, and look at the item parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group, and describing them qualitatively.
Next, we use these item parameters to systematically prune biased items from the CDI:WS, and examine how the estimated size of demographic differences change as we prune more items.
<!-- We identify a set of suspect items to eliminate (or in the future, replace), and provide updated estimates of the effect size of these demographic variables, were the biased items to be eliminated.-->
Finally, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.
Based on these analyses, we provide recommendations for next steps to be taken to identify and potentially replace items on the CDI showing demographic bias.

@@

### Sex
For sex, the median difficulty difference (male-female) was `r round(median(mm_sex$d_diff),2)` (M=`r round(mean(mm_sex$d_diff),2)`, sd=`r round(sd(mm_sex$d_diff),2)`), with `r length(which(mm_sex$d_diff>0))`/680 items being easier for females than males.
The fact that the bulk of this distribution favors females shows that the female language advantage is pervasive across the CDI:WS, and suggests that it is likely to reflect more than the effect of a small number of biased words [cf. @frank2021].
<!-- be a real ability advantage, unless the bulk of these items are actually spoken more often to female children than to male children. -->

@@

@@

```{r}
# References will be generated automatically by Pandoc and included here.
# The following code is some latex to format the bibliography. Do not remove it.
```

\setlength{\parindent}{-0.1in} 
\setlength{\leftskip}{0.125in}
\noindent

>>>>

---
title             : "Estimating demographic bias on tests of children's early vocabulary"
shorttitle        : "Demographic bias on children's vocabulary tests"

author: 
  - name          : "George Kachergis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "450 Jane Stanford Way"
    email         : "gkacherg@stanford.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
      - "Supervision"
  - name          : "Nathan Francis"
    affiliation   : "1"
    role:
      - "Writing - Review & Editing"
  - name          : "Michael C. Frank"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
      - "Supervision"

affiliation:
  - id            : "1"
    institution   : "Stanford University"
#  - id            : "2"
#    institution   : "Konstanz Business School"

authornote: |
  Department of Psychology

abstract: |
  Children's early language skill has been linked to later educational outcomes, making
  it important to measure early language accurately. Parent-reported instruments such as the 
  Communicative Development Inventories (CDIs) have been shown to provide reliable and valid
  measures of children's aggregate early language skill. However, CDIs contain
  hundreds of vocabulary items, some of which may not be heard (and thus learned) equally
  often by children of varying backgrounds. This study used a database of American English 
  CDIs to identify words demonstrating strong bias for particular demographic groups of children, on 
  dimensions of sex (male vs. female), race (white vs. non-white), and maternal education 
  (high vs. low). For each dimension, we identified dozens of strongly biased items, and showed that 
  eliminating even some of these items can reduce the magnitude of differences between groups. 
  Additionally, we investigated how well the relative frequency of words spoken to young girls vs.
  boys predicted sex-based word learning bias, and discuss possible sources of demographic
  differences in early word learning.

keywords          : "language acquisition; word learning; measuring instrument bias; demographic bias"
wordcount         : "4008"

bibliography      : "library.bib"

floatsintext      : no
linenumbers       : yes
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
r_refs("library.bib")
```

@@

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

@@

We focus here on the MacArthur-Bates Communicative Development Inventories [CDIs; @fenson1994; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
CDIs have shown good predictive validity [e.g., @fenson1994; @bornstein2012stability; @duff2015infant].
Our primary focus is the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.[^1]
For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.
Children's total vocabulary score on the CDI:WS is tightly correlated with other facets of early language (e.g., grammatical competence and gesture), suggesting that the language system is "tightly woven" [@frank2021].
Due to these desirable properties, CDIs have been adapted to dozens of languages, and a central repository of CDI data contributed from all over the world has been created [Wordbank; @frank2017; @frank2021].

[^1]: The CDI: Words and Gestures form includes a subset of the CDI:WS items and targets children 12 to 18 months of age, measuring both comprehension and production.

@@

The outline of this paper is as follows.
First, we introduce the Wordbank data and the IRT model, and examine the overall size of demographic differences in early word learning based on the full CDI:WS.
We then fit the IRT model to both groups for each demographic factor, and look at the item parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group, and describing them qualitatively.
Next, we use these item parameters to systematically prune biased items from the CDI:WS, and examine how the estimated size of demographic differences change as we prune more items.
Finally, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.
Based on these analyses, we provide recommendations for next steps to be taken to identify and potentially replace items on the CDI showing demographic bias.

@@

### Sex
For sex, the median difficulty difference (male-female) was `r round(median(mm_sex$d_diff),2)` (M=`r round(mean(mm_sex$d_diff),2)`, sd=`r round(sd(mm_sex$d_diff),2)`), with `r length(which(mm_sex$d_diff>0))`/680 items being easier for females than males.
The fact that the bulk of this distribution favors females shows that the female language advantage is pervasive across the CDI:WS, and suggests that it is likely to reflect more than the effect of a small number of biased words [cf. @frank2021].

@@

\newpage

@@

::: {#refs custom-style="Bibliography"}
:::
